Tackling discrepancies in freshwater invertebrate trait databases: Harmonising across continents and aggregating taxonomic resolution

### 2021-02-20 Paper/Submission/Main_Document_Latex.tex
20.02.: Created submission files

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\usepackage{float}
\usepackage[table,xcdraw]{xcolor}
\usepackage{color, colortbl}
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{tikz}
\usepackage{rotating}
\usepackage{caption}
\usepackage{authblk}
\usepackage{csquotes}
\graphicspath{{Figures/}}
\usepackage{setspace}
\usepackage{rotating}
\usepackage{geometry}
\usepackage{array}
\usepackage{lscape}
\usepackage{longtable}
\usepackage{etoolbox}
\usepackage{hhline}
\usepackage{lmodern}
% \usepackage[style = apa]{biblatex}
\usepackage[
backend = biber,
natbib,
citestyle = authoryear,
bibstyle = apa6,
maxcitenames = 2, 
maxbibnames = 99, 
uniquename=false,
uniquelist=false
%sorting = none % sort by name year title
]{biblatex}
\addbibresource{Ref_invertebrate_DB.bib}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% From freshwater biology:
% List all sources in the reference list alphabetically by name. In text citations should follow the author-date method. This means that the author's last name and the year of publication for the source should appear in the text, for example, (Jones, 1998), and a complete reference should appear in the reference list at the end of the paper.

% References are styled according to the sixth edition of the Publication Manual of the American Psychological Association. A sample of the most common entries in reference lists appears below. Please note that for journal articles, issue numbers are not included unless each issue in the volume begins with page one.

% Article: 
% One author: Fawcett, T. (2006). An introduction to ROC analysis. Pattern Recognition Letters, 27(8), 861–874. DOI: 10.1016/j.patrec.2005.10.010.

% 2 to 7 authors: Daley, C. E., & Nagle, R. J. (1996). Relevance of WISC-III Indicators for assessment of learning disabilities. Journal of Psychoeducational Assessment, 14 (4), 320–333.

% More than 7 authors: Rutter, M., Caspi, A., Fergusson, D., Horwood, L. J., Goodman, R., Maughan, B., … Carroll, J. (2004). Sex differences in developmental reading disability: New findings from 4 epidemiological studies. Journal of the American Medical Association, 291(16), 2007–2012. DOI: 10.1001/jama.291.16.2007

% In press or forthcoming: van Bergen, E., de Jong, P. F., Maassen, B., Krikhaar, E., Plakas, A., & van der Leij, A. (in press). IQ of four-year-olds who go on to develop dyslexia. Journal of Learning Disabilities. DOI: 10.1177/0022219413479673
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

1. Use of invertebrate traits rather than species composition may facilitate large-scale comparisons of community structure and responses to disturbance in freshwater ecology because the same traits can potentially occur everywhere.  In recent years, comprehensive invertebrate trait databases have been established at different scales (e.g. regions, continents). The wide availability of invertebrate trait data supports trait-based studies, especially at large scales. However, a number of data-related issues complicate the use of invertebrate traits for ecological studies. For example, standardised definitions for freshwater invertebrate traits are lacking, which impedes comparisons across regions. Moreover, it is uncertain how harmonising varying trait definitions among databases might influence identification of trait-environment relationships. In addition, taxonomic resolution between taxonomic composition data and trait data can differ, making aggregation of traits necessary. At present, it is unknown how different trait aggregation approaches compare with expert-assigned trait affinities. In this study we aimed to identify discrepancies in trait definitions across freshwater invertebrate databases and to compare trait aggregation methods to identify how phylogenetic structure and trait variability influence results from different aggregation approaches.

@@

3. We found that among trait databases, trait definitions often differed because varying numbers of traits were used to describe groups of related traits (e.g., respiration traits) and the focus of descriptions for groups of related traits also varied (e.g., for feeding mode some databases focused on the food source, whereas others focused on mouthpart morphology). The coding to describe traits (binary, fuzzy) also varied among databases. Our comparison of different aggregation methods showed that family-level aggregated and expert-assigned traits were similar, especially when aggregation was based on the median, and that trait coding method, uneven taxonomic hierarchical structure, and high trait variability can influence aggregated trait affinities. We further showed that harmonised and aggregated data performed similarly to not aggregated and harmonised data from the case study in identifying trait–environment relationships.

@@

Invertebrate traits have been increasingly used in freshwater ecology, e.g. by relating macroinvertebrate trait composition to environmental factors such as climate change or salinisation (\cite{bhowmik_large_2015, poff_developing_2010, szocs_effects_2014}). In the last decades, freshwater ecologists have compiled comprehensive invertebrate trait databases, which typically include data for a single region or continent (\cite{kefford_integrated_2020, Philips_and_Smith_NZ_DB_2018, schmidt-kloiber_www.freshwaterecology.info_2015, tomanova_trophic_2006, ussegliopolatera_biological_2000, vieira_database_nodate}). The availability of invertebrate trait data from different continents enables comparisons of regional trait variation and allows for testing the consistency of trait structure–environmental factor relationships across both small and large spatial extents. To date, such analyses have been carried out mostly within continents, using information from one or two trait databases. For example, \citet{bonada_taxonomic_2007} compared trait composition for Mediterranean and temperate regions in Europe based on traits from \citet{ussegliopolatera_biological_2000} (typically referred to as the Tachet database), \citet{poff_developing_2010} characterized trait composition across sites in the Western USA based on traits from \citet{poff_functional_2006}, and \citet{botwe_effects_2018} used trait definitions from \citet{poff_functional_2006} and \citet{schafer_trait_2011} to test for effects of salinity on invertebrate traits across different sites in South Australia. Analyses that synthesize invertebrate trait information from more than two different continents are rare, but see \citet{brown_functional_2018} and \citet{statzner_reproductive_1997}.

The variability and inconsistencies of information in freshwater invertebrate trait databases, besides the diversity of taxa across regions, is likely a major reason for the lack of comparisons across continents. Several challenges currently prevent use of invertebrate trait data from different established databases in ecological studies. One challenge is inconsistency in terminology. In this study, we follow the terminology proposed by \citet{schmera_proposed_2015}, in which a grouping feature (e.g. feeding mode) refers to a set of related organismal traits (e.g. predator, shredder, etc.) that together define an ecological property that is shared among individuals or species. Thus, to use consistent terminology in this study, we use the term grouping feature in place of the term trait, and the term trait instead of trait state, modality or trait category.

Another challenge to using data from multiple continents is that invertebrate trait databases include different grouping features and related traits. To harmonise grouping features from different regions, or to structure them in such a way as to facilitate cross-database comparison, commonly accepted and unambiguous trait definitions are required (\cite{schneider_towards_2019}). Ideally, the same traits and grouping features would be reported across databases or standardised terminology would exist to easily harmonise them; however, standardised terminology of trait definitions is lacking and poor metadata quality in many trait databases make, harmonisation of grouping features difficult (\cite{baird_toward_2011, kissling_towards_2018}). To our knowledge, only \citet{brown_functional_2018} harmonised grouping features from more than two geographically distant invertebrate trait databases, for a limited set of grouping features (8) and taxa (112), in a study on the influence of decreasing glacier cover on the functional diversity of invertebrate assemblages.

@@

Finally, discrepancies in the taxonomic resolutions (e.g., species, genus, or family) used when comparing among trait databases, or when linking counted taxonomic data to trait databases, presents another challenge. When taxonomic composition or data in on trait database are at a more precise taxonomic level than other trait data (e.g. observations at species-level and trait data at genus-level) trait information of the less precise taxonomic level is often assigned (e.g. \cite{szocs_effects_2014, vos_taxonomic_2017}). If trait information is only available at more precise taxonomic levels than for taxonomic composition, traits are aggregated to a less precise taxonomic level (e.g. \cite{aspin_extreme_2019, piliere_a._f._h._importance_2016, poff_functional_2006, szocs_effects_2014}). To date, studies have used different methods of trait aggregation, e.g. the mean (\cite{magliozzi_functional_2019}), median (\cite{szocs_effects_2014}) or mode (\cite{piliere_a._f._h._importance_2016}), but studies on how and to what extent different trait aggregation methods influence trait-based analyses are missing.

These challenges motivated us to assess the characteristics and comparability of different freshwater invertebrate trait databases with the goal of facilitating future syntheses. We aimed to identify methods of improving the comparability of trait data across databases, despite differences in trait definitions and the taxonomic level of data in three ways: (1) by describing discrepancies in trait definitions between trait databases from Europe, North America, Australia, and New Zealand; (2) by harmonising trait definitions across databases to provide consistent trait grouping features for comparison across geographic locations; and (3) by comparing the influence of different trait aggregation methods and grouping-feature harmonisation on the detection of trait–environment relationships.

@@

To address our study objectives, we took a multi-prong approach to describing, combining, and assessing multiple sources of freshwater invertebrate data that differed in grouping-feature definitions and taxonomic resolution. Discrepancies in trait definitions served as a starting point to develop four novel invertebrate trait datasets by harmonising seven grouping features present in the existing trait databases. The harmonised trait datasets were used to evaluate the influence of different trait aggregation approaches on the identification of trait–environment relationships, our third study objective, in two ways by (1) comparing family-level aggregated trait affinities with family-level trait affinities assigned by experts for two of the established trait datasets where such data were available and (2) comparing the influence of taxonomic phylogeny and trait variability on the outcomes of the different trait aggregation approaches for simulated data. Finally, to further investigate how harmonising and aggregating trait data can modify the outcomes of trait–environment relationship characterisation, we used data from one of the newly established datasets with harmonised and aggregated traits to repeat an analysis of salinisation effects on biological traits (\cite{szocs_effects_2014}) and compared our results with those of the original study.

@@

We established four datasets, one for each of the four geographic regions, by harmonising the traits of seven grouping features from the trait databases. We selected traits that were available in all databases, are commonly applied in trait-based ecological studies, and describe different aspects of the biology of a species: life history (Voltinism), morphology (Respiration, Body form, Size), ecology (Locomotion, Feeding mode) and reproduction (Oviposition). We omitted ecological traits that describe habitat preferences (e.g. temperature preference) because these traits are missing in the New Zealand trait database. Grouping features were classified differently across the databases, so we harmonised 7 grouping features that collectively included 26 traits (Table \ref{tab:traits_harmonisation}). To do so we combined similar traits into one trait (e.g. crawlers and sprawlers into crawlers) by assigning for each taxon for the new combined trait the highest affinity score of those originally assigned. Next, we selected taxa for inclusion in our datasets and normalised the trait affinity scores for each taxon. We first omitted all taxa with a lower taxonomic precision than family level and consolidated duplicate taxa that appeared within a database, by either applying the median for fuzzy coded traits, or the maximum for binary traits. We then used either fuzzy coding or, when unavailable, binary coding to create the final harmonised datasets. Fuzzy codes are reported with different ranges in the trait databases (e.g., freshwaterecology.info: 0–10, Tachet database: 0–3 or 0–5). We normalised these to a range between 0 and 1 and converted trait affinities to proportions. In cases where binary coding was needed, we converted categorical and continuous traits into binary traits with values of 1 and 0, which, therefore, had the same range as fuzzy coded traits. For example, in the  freshwaterecology.info database, voltinism traits are associated with different faunistic regions, and we substituted a value of 1 for entries such as “arctic” or “boreal” % and 0 if no region was given. 
We assumed that values of 1 and 0 for binary traits corresponded to the highest and lowest affinity for a particular trait.

\begin{longtable}{m{2.5cm}|m{4cm}|m{7cm}}

@@

To examine the influence of phylogeny and trait variability on the outcomes of the different trait aggregation methods, we created three hypothetical scenarios. We simulated three different families, each containing 25 total species but with different phylogenetic structures. The three families consisted of (1) a family with an equal number of genera and species (five genera each with five species each), denoted as \textit{sim\_base}; (2) a family in which one genus had a much larger number of species than the other four genera (1 genus with 13 species, 4 genera with 3 species each), denoted as \textit{sim\_extreme}; (3) a family in which all genera had a different number of species (8, 2, 7, 3, 5), denoted as \textit{sim\_variation}. We assigned a hypothetical grouping feature with three traits (T1, T2, and T3) to each scenario. We then simulated the 25 affinities, one per species, for each trait, by sampling from a truncated normal distribution bound by 0 and 1 and with a mean value of 0.5. To simulate different levels of trait variability, we repeated the sampling 100 times for each of the 5 standard deviations (0.2, 0.4, 0.6, 0.8, and 1), resulting in 12,500 simulated trait affinities for each simulated trait ($25$ species per family $\times$ 5 levels of trait variability $\times$ 100 replicates). We converted simulated trait affinities to proportions, as was done during trait database processing, and assigned the 12,500 simulated affinities per trait to each of the three family scenarios. We then applied the five trait aggregation methods described above to each simulated dataset. We compared the resulting ranges of aggregated trait affinities between levels of trait variability and taxonomic hierarchical scenario as well as the differences in trait affinities obtained by each aggregation method.

\begin{figure}
  \centering
  % \includegraphics[width=22cm, height=28cm]{Flowchart_methods.pdf}
  \subfile{Flowchart/Flowchart_methods.tex}
  % \label{fig:data_proc_overview}
\end{figure}
\newpage
\begin{center}
\captionof{figure}{Data processing steps of the selected traits. Intermediate (grey) and main (orange) steps of data preparation are depicted. The dashed bottom box illustrates the different trait aggregation methods (hypothetical data in the upper left corner). The aggregation methods (purple) and intermediate steps of the aggregation methods (pink) are displayed. Abbreviations: EU, Europe; NOA, North America; AUS, Australia; NZ, New Zealand.}
\label{fig:data_proc_overview}
\end{center}

@@

We re-analysed the \citet{szocs_effects_2014} data by performing a RDA with harmonised grouping features from the dataset for Europe and with traits that we aggregated to family level with the five aggregation methods described above. We used all 21 grouping features from the original study, but replaced 6 with harmonised grouping features from the established European trait dataset, namely size, feeding mode, locomotion, oviposition, respiration, and voltinism.
We compared the results of the original study’s RDA with the results of the RDA from the harmonised grouping features. Specifically, trait composition, expressed as community weighted mean (CWM) traits, was ordinated along an electrical conductivity gradient. We compared the species scores obtained from the RDA (i.e., the coordinates of the tips of the vectors representing the CWM traits in the bi- or triplots). Following the original study, we used the Mahalanobis distance measure and identified traits associated with higher or lower salinity based on their distance to the ordination axis median. We considered CWM traits with a Mahalanobis distance greater than the 97.5 \%-quantile of the Chi-square distribution (5.02) to have responded to either lower or higher salinity. Additionally, to test fo the effect of different aggregation methods on inferences about CWM trait–salinisation relationships, we repeated the RDA five times, each with family-level aggregated trait values assigned to corresponding taxa from \citet{szocs_effects_2014}.

@@

All data processing and RDA analyses were carried out using R version 3.6.1 (\cite{cite_R}). RDA was computed using the vegan package (\cite{cite_vegan}). Raw data and the R code for data processing and grouping feature harmonisation are located in the Github repository: \url{https://github.com/KunzstLD/Invertebrate_traits}. Scripts and data to reproduce the trait aggregation, comparison of aggregated and assigned traits, and the RDA analysis are located in the Github repository \url{https://github.com/KunzstLD/Trait-aggregation}.

@@

\begin{longtable}{m{2.3cm}|m{3.3cm}|m{1.8cm}|m{1.4cm}|m{1.4cm}|m{4.4cm}|m{2cm}}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=16.5cm, height=10cm]{Overview_sim_results.png}
  \caption{Ranges of aggregated trait affinities for the three examples of taxonomic hierarchies and simulated levels of trait variability. Shown are the results only for one simulated trait (T1). Similar results were obtained for the other simulated traits (Figure \ref{fig:overview_sim_results_T2_T3}). Boxplots depict results for 100 replicated simulations of each trait aggregation method. The boxplot depicts the median and encompasses the 25\textsuperscript{th} and 75\textsuperscript{th} percentile. Horizontal black lines depict the median. Whiskers extend to the largest and smallest value respectively no further than 1.5 $*$ the inter-quartile range. Outliers beyond the end of the whiskers are plotted as grey dots. Trait aggregation methods are in order of least to greatest produced ranges to improve visual inspection.}
  \label{fig:overview_sim_results}
\end{figure}

@@

% From Freshwater biology: 
% This should highlight the significance of the results and place them in the context of other work. It should not include references to figures or tables cited in the Results – check that all of your Results are presented in the Results section. Figures (e.g. photographs, conceptual diagrams) or tables presented specifically as part of the Discussion are welcome and may be cited there.

@@

Synthesizing trait information from multiple trait databases is crucial for developing the full potential of trait-based approaches, such as for studying community trait responses to environmental gradients across geographic regions. Others have noted the lack of standardised trait terminology in freshwater ecology (\cite{baird_toward_2011, brink_traits-based_2011}), but in this study we show for the first time the discrepancies that exist across some commonly used grouping features. Furthermore, we offer insight into the need for, and process of, harmonising invertebrate trait information from trait databases of different regions. Some of the challenges we found were that grouping features were resolved differently into traits in different databases and that various codings (fuzzy, categorical, binary) were used to describe traits. In addition, the same traits were sometimes defined differently, requiring expert knowledge to reclassify these traits (e.g. the trait piercer). Missing information was another challenge. We aimed to use most of the available invertebrate trait information for different regions to establish harmonised grouping feature datasets, but the availability of trait information varied strongly across grouping features and databases. For instance, extensive information was available for grouping features that are often used in trait–environment studies, such as feeding mode and respiration, but there was surprisingly little information on body form, a trait that is relatively easy to characterise.

@@

As ecological knowledge increases, and with the help of new methods such as computer vision for species and trait identification (\cite{hoye_deep_2020}), more observed taxa will be identified to the species level and trait data will be available at the species level in the future, reducing the need for trait aggregation. However, we showed that trait aggregation is often in agreement with expert assignments, especially through aggregation approaches that use the median. Based on this analysis, we cannot claim that a specific aggregation method is objectively the best, and except for cases where traits are highly variable and phylogenetic structure is highly uneven, the choice of aggregation method does not appear to be influential to the outcome. Weighting trait information based on the taxa present in the database used does not seem to have much influence except when the taxonomic hierarchical structure is highly uneven. However, we recommend that the type of coding used to describe traits (binary, fuzzy) should be considered, because median and mean aggregations can yield very different results for binary-coded traits. We also suggest that a measure of trait dispersion should be reported to indicate the uncertainty of the aggregated estimate. Use of carefully harmonised and integrated trait data as described here should  enhance our ability to study a wide range of large-scale ecological questions with greater rigor than currently exists.

>>>>

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\usepackage{float}
\usepackage[table,xcdraw]{xcolor}
\usepackage{color, colortbl}
\usepackage{longtable}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{tikz}
\usepackage{rotating}
\usepackage{caption}
\usepackage{authblk}
\usepackage{csquotes}
\graphicspath{{Figures/}}
\usepackage{setspace}
\usepackage{rotating}
\usepackage{geometry}
\usepackage{array}
\usepackage{lscape}
\usepackage{longtable}
\usepackage{etoolbox}
\usepackage{hhline}
\usepackage{lmodern}
\usepackage[
backend = biber,
natbib,
citestyle = authoryear,
bibstyle = apa6,
maxcitenames = 2, 
maxbibnames = 99, 
uniquename=false,
uniquelist=false
%sorting = none % sort by name year title
]{biblatex}
\addbibresource{Ref_invertebrate_DB.bib}

@@

1. Use of invertebrate traits rather than species composition may facilitate large-scale comparisons of community structure and responses to disturbance in freshwater ecology because the same traits can potentially occur everywhere.  In recent years, comprehensive invertebrate trait databases have been established at different scales (e.g. regions, continents). The wide availability of invertebrate trait data supports trait-based studies, especially at large scales. However, a number of data-related issues complicate the use of invertebrate traits for ecological studies. For example, standardised definitions for freshwater invertebrate traits are lacking, which impedes comparisons across regions. Moreover, it is uncertain how harmonising varying trait definitions among databases might influence identification of trait-environment relationships. In addition, taxonomic resolution between taxonomic composition data and trait data can differ, making aggregation of traits necessary. At present, it is unknown how different trait aggregation approaches compare with expert-assigned trait affinities. In this study we aimed to identify discrepancies in trait definitions across freshwater invertebrate databases and to compare trait aggregation methods to identify how phylogenetic structure and trait variability influence results from different aggregation methods.

@@

3. We found that among trait databases, trait definitions often differed because varying numbers of traits were used to describe groups of related traits (e.g., respiration traits) and the focus of descriptions for groups of related traits also varied (e.g., for feeding mode some databases focused on the food source, whereas others focused on mouthpart morphology). The coding to describe traits (binary, fuzzy) also varied among databases. Our comparison of different aggregation methods showed that family-level aggregated and expert-assigned traits were similar, especially when aggregation was based on the median, and that trait coding method, uneven phylogenetic structure, and high trait variability can influence aggregated trait affinities. We further showed that harmonised and aggregated data performed similarly to not aggregated and harmonised data from the case study in identifying trait–environment relationships.

@@

Invertebrate traits have been increasingly used in freshwater ecology, e.g. by relating macroinvertebrate trait composition to environmental factors such as climate change or salinisation (\cite{bhowmik_large_2015, poff_developing_2010, szocs_effects_2014}). In the last decades, freshwater ecologists have compiled comprehensive invertebrate trait databases, which typically include data for a single region or continent (\cite{kefford_integrated_2020, Philips_and_Smith_NZ_DB_2018, schmidt-kloiber_www.freshwaterecology.info_2015, tomanova_trophic_2006, ussegliopolatera_biological_2000, vieira_database_nodate}). The availability of invertebrate trait data from different continents enables comparisons of regional trait variation and allows for testing the consistency of trait structure–environmental factor relationships across both small and large spatial extents. To date, such analyses have been carried out mostly within continents, using information from one or two trait databases. For example, \citet{bonada_taxonomic_2007} compared trait composition for Mediterranean and temperate regions in Europe based on traits from \citet{ussegliopolatera_biological_2000} (typically referred to as the Tachet database), \citet{poff_developing_2010} characterised trait composition across sites in the Western USA based on traits from \citet{poff_functional_2006}, and \citet{botwe_effects_2018} used trait definitions from \citet{poff_functional_2006} and \citet{schafer_trait_2011} to test for effects of salinity on invertebrate traits across different sites in South Australia. Analyses that synthesise invertebrate trait information from more than two different continents are rare, but see \citet{brown_functional_2018} and \citet{statzner_reproductive_1997}.

The variability and inconsistencies of information in freshwater invertebrate trait databases, besides the diversity of taxa across regions, is likely a major reason for the lack of comparisons across continents. Several challenges currently prevent the use of invertebrate trait data from different established databases in ecological studies. One challenge is inconsistency in terminology. In this study, we follow the terminology proposed by \citet{schmera_proposed_2015}, in which a grouping feature (e.g. feeding mode) refers to a set of related traits (e.g. predator, shredder, etc.) that together define an ecological property that is shared among individuals or species. Thus, to use consistent terminology in this study, we use the term grouping feature in place of the term trait, and the term trait instead of trait state, modality or trait category.

Another challenge to using data from multiple continents is that invertebrate trait databases include different grouping features and related traits. To harmonise grouping features from different regions, or to structure them in such a way as to facilitate cross-database comparison, commonly accepted and unambiguous trait definitions are required (\cite{schneider_towards_2019}). Ideally, the same traits and grouping features would be reported across databases or standardised terminology would exist to easily harmonise them; however, standardised terminology of trait definitions is lacking and poor metadata quality in many trait databases make harmonisation of grouping features difficult (\cite{baird_toward_2011, kissling_towards_2018}). To our knowledge, only \citet{brown_functional_2018} harmonised grouping features from more than two geographically distant invertebrate trait databases, for a limited set of grouping features (8) and taxa (112), in a study on the influence of decreasing glacier cover on the functional diversity of invertebrate assemblages.

@@

Finally, discrepancies in the taxonomic resolutions (e.g., species, genus, or family) used when comparing among trait databases, or when linking observed taxonomic data to trait databases, presents another challenge. When taxonomic composition or data in on trait database are at a more precise taxonomic level than other trait data (e.g. observations at species-level and trait data at genus-level) trait information of the less precise taxonomic level is often assigned (e.g. \cite{szocs_effects_2014, vos_taxonomic_2017}). If trait information is only available at more precise taxonomic levels than for taxonomic composition, traits are aggregated to a less precise taxonomic level (e.g. \cite{aspin_extreme_2019, piliere_a._f._h._importance_2016, poff_functional_2006, szocs_effects_2014}). To date, studies have used different methods of trait aggregation, e.g. the mean (\cite{magliozzi_functional_2019}), median (\cite{szocs_effects_2014}) or mode (\cite{piliere_a._f._h._importance_2016}), but studies on how and to what extent different trait aggregation methods influence trait-based analyses are missing.

These challenges motivated us to assess the characteristics and comparability of different freshwater invertebrate trait databases with the goal of facilitating future syntheses. We aimed to identify methods of improving the comparability of trait data across databases, despite differences in trait definitions and the taxonomic level of data in three ways: (1) by describing discrepancies in trait definitions between trait databases from Europe, North America, Australia, and New Zealand; (2) by harmonising trait definitions across databases to provide consistent grouping features for comparison across geographic locations; and (3) by comparing the influence of different trait aggregation methods and grouping-feature harmonisation on the detection of trait–environment relationships.

@@

To address our study objectives, we took a multi-prong approach to describing, combining, and assessing multiple sources of freshwater invertebrate data that differed in grouping-feature definitions and taxonomic resolution. Discrepancies in trait definitions served as a starting point to develop four novel invertebrate trait datasets by harmonising seven grouping features present in the existing trait databases. The harmonised trait datasets were used to evaluate the influence of different trait aggregation methods on the identification of trait–environment relationships, our third study objective, in two ways by (1) comparing family-level aggregated trait affinities with family-level trait affinities assigned by experts for two of the established trait datasets where such data were available and (2) comparing the influence of phylogenetic structure and trait variability on the outcomes of the different trait aggregation approaches for simulated data. Finally, to further investigate how harmonising and aggregating trait data can modify the outcomes of trait–environment relationship characterisation, we used data from one of the newly established datasets with harmonised and aggregated traits to repeat an analysis of salinisation effects on biological traits (\cite{szocs_effects_2014}) and compared our results with those of the original study.

@@

We established four datasets, one for each of the four geographic regions, by harmonising the traits of seven grouping features from the trait databases. We selected traits that were available in all databases, are commonly applied in trait-based ecological studies, and describe different aspects of the biology of a species: life history (Voltinism), morphology (Respiration, Body form, Size), ecology (Locomotion, Feeding mode) and reproduction (Oviposition). We omitted ecological traits that describe habitat preferences (e.g. temperature preference) because these traits are missing in the New Zealand trait database. Grouping features were classified differently across the databases, so we harmonised 7 grouping features that collectively included 26 traits (Table \ref{tab:traits_harmonisation}). To do so we combined similar traits into one trait (e.g. crawlers and sprawlers into crawlers) by assigning for each taxon for the new combined trait the highest affinity score of those originally assigned. Next, we selected taxa for inclusion in our datasets and normalised the trait affinity scores for each taxon. We first omitted all taxa with a lower taxonomic precision than family level and consolidated duplicate taxa that appeared within a database, by either applying the median for fuzzy coded traits, or the maximum for binary traits. We then used either fuzzy coding or, when unavailable, binary coding to create the final harmonised datasets. Fuzzy codes are reported with different ranges in the trait databases (e.g., freshwaterecology.info: 0–10, Tachet database: 0–3 or 0–5). We normalised these to a range between 0 and 1 and converted trait affinities to proportions. In cases where binary coding was needed, we converted categorical and continuous traits into binary traits with values of 1 and 0, which, therefore, had the same range as fuzzy coded traits. For example, in the  freshwaterecology.info database, voltinism traits are associated with different faunistic regions, and we substituted a value of 1 for entries such as “arctic” or “boreal”. We assumed that values of 1 and 0 for binary traits corresponded to the highest and lowest affinity for a particular trait.

\begin{longtable}{m{2.5cm}|m{4cm}|m{7.5cm}}

@@

To examine the influence of phylogeny and trait variability on the outcomes of the different trait aggregation methods, we created three hypothetical scenarios. We simulated three different families, each containing 25 total species but with different phylogenetic structures. The three families consisted of (1) a family with an equal number of genera and species (five genera each with five species each), denoted as \textit{sim\_base}; (2) a family in which one genus had a much larger number of species than the other four genera (1 genus with 13 species, 4 genera with 3 species each), denoted as \textit{sim\_extreme}; (3) a family in which all genera had a different number of species (8, 2, 7, 3, 5), denoted as \textit{sim\_variation}. We assigned a hypothetical grouping feature with three traits (T1, T2, and T3) to each scenario. We then simulated the 25 affinities, one per species, for each trait, by sampling from a truncated normal distribution bound by 0 and 1 and with a mean value of 0.5. To simulate different levels of trait variability, we repeated the sampling 100 times for each of the 5 standard deviations (0.2, 0.4, 0.6, 0.8, and 1), resulting in 12,500 simulated trait affinities for each simulated trait ($25$ species per family $\times$ 5 levels of trait variability $\times$ 100 replicates). We converted simulated trait affinities to proportions, as was done during trait database processing, and assigned the 12,500 simulated affinities per trait to each of the three family scenarios. We then applied the five trait aggregation methods described above to each simulated dataset. We compared the resulting ranges of aggregated trait affinities between levels of trait variability and phylogenetic scenario as well as the differences in trait affinities obtained by each aggregation method.

\begin{figure}
  \centering
  \subfile{Flowchart/Flowchart_methods.tex}
\end{figure}
\newpage
\begin{center}
\captionof{figure}{Data processing steps of the selected traits. Intermediate (grey) and main (orange) steps of data preparation are depicted. The dashed bottom box illustrates the different trait aggregation methods (hypothetical data in the upper left corner). The aggregation methods (purple) and intermediate steps of the aggregation methods (pink) are displayed. Abbreviations: EU, Europe; NOA, North America; AUS, Australia; NZ, New Zealand.}
\label{fig:data_proc_overview}
\end{center}

@@

We re-analysed the \citet{szocs_effects_2014} data by performing a RDA with harmonised grouping features from the dataset for Europe and with traits that we aggregated to family level with the five aggregation methods described above. We used all 21 grouping features from the original study, but replaced 6 with harmonised grouping features from the established European trait dataset, namely size, feeding mode, locomotion, oviposition, respiration, and voltinism.
We compared the results of the original study’s RDA with the results of the RDA from the harmonised grouping features. Specifically, trait composition, expressed as community weighted mean (CWM) traits, was ordinated along an electrical conductivity gradient. We compared the species scores obtained from the RDA (i.e., the coordinates of the tips of the vectors representing the CWM traits in the bi- or triplots). Following the original study, we used the Mahalanobis distance measure and identified traits associated with higher or lower salinity based on their distance to the ordination axis median. We considered CWM traits with a Mahalanobis distance greater than the 97.5 \%-quantile of the Chi-square distribution (5.02) to have responded to either lower or higher salinity. Additionally, to test for the effect of different aggregation methods on inferences about CWM trait–salinisation relationships, we repeated the RDA five times, each with family-level aggregated trait values assigned to corresponding taxa from \citet{szocs_effects_2014}.

@@

All data processing and RDA analyses were carried out using R version 3.6.1 (\cite{cite_R}). RDA was computed using the vegan package (\cite{cite_vegan}). Raw data and the R code for data processing and grouping feature harmonisation are located in the Github repository: \url{https://github.com/KunzstLD/Invertebrate_traits}. Scripts and data to reproduce the trait aggregation, comparison of aggregated and assigned traits, and the RDA analysis are located in the Github repository: \url{https://github.com/KunzstLD/Trait-aggregation}.

@@

\begin{longtable}{m{2.5cm}|m{3.5cm}|m{2cm}|m{1.5cm}|m{1.5cm}|m{4.8cm}|m{2cm}}

@@

\begin{figure}[H]
  \centering
  \includegraphics[width=16.5cm, height=10cm]{Overview_sim_results.png}
  \caption{Ranges of aggregated trait affinities for the three examples of taxonomic hierarchies and simulated levels of trait variability. Shown are the results only for one simulated trait (T1). Similar results were obtained for the other simulated traits (Figure \ref{fig:overview_sim_results_T2_T3}). Boxplots depict results for 100 replicated simulations of each trait aggregation method. The boxplot depicts the median and encompasses the 25\textsuperscript{th} and 75\textsuperscript{th} percentile. Horizontal black lines depict the median. Whiskers extend to the largest and smallest value respectively no further than 1.5 $\times$ the inter-quartile range. Outliers beyond the end of the whiskers are plotted as grey dots. Trait aggregation methods are in order of least to greatest produced ranges to improve visual inspection.}
  \label{fig:overview_sim_results}
\end{figure}

@@

@@

Synthesising trait information from multiple trait databases is crucial for developing the full potential of trait-based approaches, such as for studying community trait responses to environmental gradients across geographic regions. Others have noted the lack of standardised trait terminology in freshwater ecology (\cite{baird_toward_2011, brink_traits-based_2011}), but in this study we show for the first time the discrepancies that exist across some commonly used grouping features. Furthermore, we offer insight into the need for, and process of, harmonising invertebrate trait information from trait databases of different regions. Some of the challenges we found were that grouping features were resolved differently into traits in different databases and that various codings (fuzzy, categorical, binary) were used to describe traits. In addition, the same traits were sometimes defined differently, requiring expert knowledge to reclassify these traits (e.g. the trait piercer). Missing information was another challenge. We aimed to use most of the available invertebrate trait information for different regions to establish harmonised grouping feature datasets, but the availability of trait information varied strongly across grouping features and databases. For instance, extensive information was available for grouping features that are often used in trait–environment studies, such as feeding mode and respiration, but there was surprisingly little information on body form, a trait that is relatively easy to characterise.

@@

As ecological knowledge increases, and with the help of new methods such as computer vision for species and trait identification (\cite{hoye_deep_2020}), more observed taxa will be identified to the species level and trait data will be available at the species level in the future, reducing the need for trait aggregation. However, we showed that trait aggregation is often in agreement with expert assignments, especially through aggregation approaches that use the median. Based on this analysis, we cannot claim that a specific aggregation method is objectively the best, and except for cases where traits are highly variable and phylogenetic structure is highly uneven, the choice of aggregation method does not appear to be influential to the outcome. Weighting trait information based on the taxa present in the database used does not seem to have much influence except when the phylogenetic structure is highly uneven. However, we recommend that the type of coding used to describe traits (binary, fuzzy) should be considered, because median and mean aggregations can yield very different results for binary-coded traits. We also suggest that a measure of trait dispersion should be reported to indicate the uncertainty of the aggregated estimate. Use of carefully harmonised and integrated trait data as described here should enhance our ability to study a wide range of large-scale ecological questions with greater rigor than currently exists.
